---
title: Disentangling Composition and Spatial Effects
params:
  K: 20
---

```{r}
library("ggplot2")
library("dplyr")
library("reshape2")
library("tibble")
library("igraph")
library("forcats")
source("preprocessing.R")
theme_set(theme_bw() + theme(panel.grid=element_blank()))
```

```{r, message = FALSE, warning = FALSE}
data_dir <- file.path("..", "Data")
loaded_ <- load_mibi(file.path("..", "Data"))
subsample <- spatial_subsample(loaded_$tiffs, loaded_$mibi)
ims <- subsample$ims
mibi_sce <- subsample$exper
```

* get cluster the proteins, and extract compositions from each sample
* plot this against the tumor intrusion
* now calculate my own spatial statistics.
  - neighborhood entropy
  - neighborhood size
* relate summaries of these spatial statistics to composition
  - if not totally predictive, then the latent idea is worthwhile
  - otherwise, it's a surprise to the community

```{r}
cell_clusters <- kmeans(t(assay(mibi_sce)), params$K)
props <- sample_proportions(colData(mibi_sce)$SampleID, cell_clusters$cluster)
mprops <- data.frame(props)
```

```{r}
cd <- colData(mibi_sce) %>%
  as.data.frame() %>%
  tidyr::unite(scell, SampleID, cellLabelInImage, remove=FALSE) %>%
  mutate(
    cell_group = fct_lump(cell_type, prop = 0.05),
    SampleID = factor(SampleID, rownames(props))
  )

cd_samp <- cd %>%
  group_by(SampleID) %>%
  summarise_all(function(x) { x[1] }) %>%
  mutate(SampleID = factor(SampleID, rownames(props)))

cluster_ids <- setNames(cell_clusters$cluster, cd$scell)
```

These are example sample-level cluster compositions, based entirely on antigen
information (ignoring spatial information).
```{r}
ggplot(mprops %>% left_join(cd_samp)) +
  geom_tile(
    aes(x = SampleID, y = cluster, fill = sqrt(Freq))
  ) +
  facet_grid(. ~ GRADE, scale="free_x", space="free_x") +
  scale_fill_gradient(low = "white", high = "black") +
  theme(legend.position = "bottom")
```


Extract local entropy and neighborhood sizes.
```{r spatial_stat, message = FALSE, warning = FALSE}
sample_names <- as.character(cd_samp$SampleID)
graphs <- list()
for (i in seq_along(sample_names)) {
  print(sprintf("graph %s/%s", i, length(sample_names)))
  poly <- polygonize(ims[[sample_names[i]]]) %>%
    filter(cellLabelInImage > 1)
  graphs[[i]] <- extract_graph(poly)
}

spatial <- list()
for (i in seq_along(sample_names)) {
  print(sprintf("spatial stats %s/%s", i, length(sample_names)))
  SG <- subgraphs(graphs[[i]])
  ptrn <- paste0("^", sample_names[i], "_")
  clusters_ <- cluster_ids[grepl(ptrn, names(cluster_ids))]
  names(clusters_) <- gsub(ptrn, "", names(clusters_))

  spatial[[sample_names[i]]] <- tibble(
    scell = paste0(sample_names[i], "_", names(V(graphs[[i]]))),
    entropy = entropies(SG, clusters_),
    avg_dists = avg_dists(SG)
  )
}
```

```{r plot_spatial}
spatial <- bind_rows(spatial, .id = "SampleID") %>%
  mutate(SampleID = factor(SampleID, levels(mprops$SampleID)))

spatial_samp <- spatial %>%
  inner_join(cd) %>%
  group_by(SampleID) %>%
  summarise(
    me = mean(entropy),
    sde = sd(entropy),
    mdist = mean(avg_dists),
    sdist = sd(avg_dists),
    .groups = "drop"
  ) %>%
  left_join(cd_samp)

spatial_cell <- spatial %>%
  tidyr::separate(scell, c("SampleID", "cellLabelInImage")) %>%
  mutate(
    cellLabelInImage = as.numeric(cellLabelInImage)
  ) %>%
  left_join(cd)
```


```{r}
## overall histograms
ggplot(spatial_cell) +
  geom_histogram(aes(x = entropy, fill = as.factor(TIL_score))) +
  scale_fill_brewer(palette = "Greens", na.value = "grey") +
  facet_grid(cell_group ~ ., scale = "free") +
  theme(
    legend.position = "bottom",
    strip.text.y = element_text(angle = 0)
  )

ggplot(spatial_cell) +
  geom_histogram(aes(x = avg_dists, fill = as.factor(TIL_score))) +
  scale_fill_brewer(palette = "Greens", na.value = "grey") +
  facet_grid(cell_group ~ ., scale = "free") +
  theme(
    legend.position = "bottom",
    strip.text.y = element_text(angle = 0)
  )

slev <- spatial_samp %>%
  select(SampleID, mdist) %>%
  arrange(mdist) %>%
  .[["SampleID"]]

ggplot(spatial_cell %>%
       mutate(SampleID = factor(SampleID, levels = slev))
       ) +
  geom_point(
    aes(x = avg_dists, y = entropy, col = cell_group),
    size = 0.5, alpha = 0.8
  ) +
  facet_wrap(~ SampleID) +
  scale_fill_brewer(palette = "Set2") +
  theme(
    legend.position = "bottom",
    strip.text.y = element_text(angle = 0),
    panel.spacing = unit(0, "cm")
  )
```

This is how each individual differs in terms of the cell-label entropy of and
average pairwise distance of their 5-nearest neighborhoods.

```{r}
## Sample level plots
ggplot(spatial_samp) +
  geom_point(
    aes(x = me, y = Survival_days_capped_2016.1.1),
    size = 4
  ) +
  geom_point(
    aes(x = me, y = Survival_days_capped_2016.1.1, col = as.factor(TIL_score)),
    size = 2
  ) +
  scale_color_brewer(palette = "Greens", na.value = "grey") +
  theme(legend.position = "bottom")

ggplot(spatial_samp) +
  geom_point(
    aes(x = mdist, y = Survival_days_capped_2016.1.1),
    size = 4
  ) +
  geom_point(
    aes(x = mdist, y = Survival_days_capped_2016.1.1, col = as.factor(TIL_score)),
    size = 2
  ) +
  scale_color_brewer(palette = "Greens", na.value = "grey") +
  theme(legend.position = "bottom")

## Cell level plots
ggplot(spatial_cell) +
  geom_jitter(
    aes(x = TIL_score, y = entropy, col = as.factor(SampleID)),
    size = 0.8, alpha = 0.8
  ) +
  theme(legend.position = "none")

ggplot(spatial_cell) +
  geom_jitter(
    aes(x = TIL_score, y = avg_dists, col = as.factor(SampleID)),
    size = 0.8, alpha = 0.8
  ) +
  theme(legend.position = "none")
```

* Can we predict properties of the sample-wise entropy distribution based on
  just the cell compositions?
* Is there something about the survival that isn't captured entirely by the
  composition information?

```{r}
spatial_samp <- spatial_samp %>%
  mutate(SampleID = factor(SampleID, levels(mprops$SampleID)))

spatial_comp <- spatial_samp %>%
  select(SampleID, me, sde, mdist, sdist, TIL_score) %>%
  left_join(mprops) %>%
  dcast(SampleID + me + sde + mdist + sdist + TIL_score ~ cluster)

x <- spatial_comp %>%
  select(matches("[0-9]+")) %>%
  as.matrix() %>%
  sqrt()
```

```{r}
y <- as.numeric(scale(spatial_comp$me))
fits <- fit_wrapper(x, y)
plot_fits(x, y, fits$glmnet, fits$rf)
```

It looks like quite a bit, but not all, of the variation in average entropy, can
be explained by cell composition. What about the average neighborhood size or
standard errors of entropy and neighborhood size?

```{r}
y <- as.numeric(scale(spatial_comp$mdist))
fits <- fit_wrapper(x, y)
plot_fits(x, y, fits$glmnet, fits$rf)

y <- as.numeric(scale(spatial_comp$sde))
fits <- fit_wrapper(x, y)
plot_fits(x, y, fits$glmnet, fits$rf)

y <- as.numeric(scale(spatial_comp$sdist))
fits <- fit_wrapper(x, y)
plot_fits(x, y, fits$glmnet, fits$rf)
```

It seems like average pairwise distance is predictable, though not quite as much
as entropy. The standard errors are not predictable using the linear model, but
seem okay using the random forest (which I find an odd result...)

What about the TIL score?

```{r}
y <- as.numeric(scale(spatial_comp$TIL_score))
keep_ix <- !is.na(y)
sum(keep_ix)

fits <- fit_wrapper(x[keep_ix, ], y[keep_ix])
plot_fits(x[keep_ix, ], y[keep_ix], fits$glmnet, fits$rf)
```

There are so many missing values in TIL score, that I'm not sure whether we're
actually learning anything meaningful here. It may be the case that the entropy
and neighborhood size features are predictable, but not the TIL score? I don't
have an explanation for how that could happen, though.

# Interpreting Models

```{r}
y <- as.numeric(scale(spatial_comp$me))
fits <- fit_wrapper(x, y)
```

First, let's study the output of the glmnet model, even though it's performance is lower.

```{r}
plot(fits$glmnet$glmnet.fit)
beta_hat <- coef(fits$glmnet$glmnet.fit)[, 10]

imp_ix <- order(abs(beta_hat), decreasing = TRUE)

mx <- data.frame(x = x[, imp_ix[1:10] - 1], y = y) %>%
  melt(id.vars = c("y"))

ggplot(mx) +
  geom_point(aes(x = value, y = y)) +
  facet_wrap(~ variable)

y_order <- order(y)
image(t(cbind(y[y_order] / 4, x[y_order, imp_ix[1:10] - 1])))
```

How can we interpret the associated clusters?

```{r}
z <- cell_clusters$centers
mcenters <- melt(z, varnames = c("cluster", "protein")) %>%
  mutate(
    protein = factor(protein, colnames(z)[hclust(dist(t(z), method="manhattan"))$order]),
    cluster = factor(cluster, hclust(dist(z, method = "manhattan"))$order),
    )

mcenters <- mcenters %>%
  left_join(data.frame(
      cluster = factor(0:20, levels = c(0, levels(mcenters$cluster))),
      beta = beta_hat
    ))

ggplot(mcenters) +
  geom_bar(
    aes(x = protein, y = value, fill = beta),
    stat = "identity"
  ) +
  scale_fill_viridis() +
  scale_y_continuous(limits = c(-2, 6), oob = scales::squish) +
  facet_wrap(~ cluster) +
theme(axis.text.x = element_text(angle = -90, hjust = 0))

ggplot(mcenters %>% filter(beta != 0)) +
  geom_bar(
    aes(x = protein, y = value, fill = beta),
    stat = "identity"
  ) +
  scale_fill_viridis() +
  scale_y_continuous(limits = c(-2, 6), oob = scales::squish) +
  facet_grid(cluster ~ .) +
  theme(axis.text.x = element_text(angle = -90, hjust = 0))
```

We can make something similar for the random forest model.


# Inspecting Residuals

# F-Statistic Analog



# 
