---
title: Disentangling Composition and Spatial Effects
params:
  K: 20
---

```{r}
library("ggplot2")
library("dplyr")
library("reshape2")
library("tibble")
library("igraph")
library("forcats")
theme_set(theme_bw() + theme(panel.grid=element_blank()))
```

```{r}
data_dir <- file.path("..", "Data")
loaded_ <- load_mibi(file.path("..", "Data"), 20)
subsample <- spatial_subsample(loaded_$tiffs, loaded_$mibi)
ims <- subsample$ims
mibi_sce <- subsample$exper
```

* get cluster the proteins, and extract compositions from each sample
* plot this against the tumor intrusion
* now calculate my own spatial statistics.
  - neighborhood entropy
  - neighborhood size
* relate summaries of these spatial statistics to composition
  - if not totally predictive, then the latent idea is worthwhile
  - otherwise, it's a surprise to the community

```{r}
cell_clusters <- kmeans(t(assay(mibi_sce)), params$K)
props <- sample_proportions(assay(mibi_sce), cell_clusters)
```

```{r}
cd <- colData(mibi_sce) %>%
  as.data.frame() %>%
  unite(scell, SampleID, cellLabelInImage, remove=FALSE) %>%
  mutate(cell_group = fct_lump(cell_type, prop = 0.05))

cd_samp <- cd %>%
  group_by(SampleID) %>%
  summarise_all(function(x) { x[1] })

cluster_ids <- setNames(cell_clusters$cluster, cd$scell)
plot(cd_samp$TIL_score + runif(nrow(cd_samp), 0, 0.1), cd_samp$Survival_days_capped_2016.1.1)
```

Extract local entropy and neighborhood sizes.
```{r spatial_stat}
sample_names <- as.character(cd_samp$SampleID)
graphs <- list()
for (i in seq_along(sample_names)) {
  print(sprintf("graph %s/%s", i, length(sample_names)))
  poly <- polygonize(ims[[sample_names[i]]]) %>%
    filter(cellLabelInImage > 1)
  graphs[[i]] <- extract_graph(poly)
}

spatial <- list()
for (i in seq_along(sample_names)) {
  print(sprintf("spatial stats %s/%s", i, length(sample_names)))
  SG <- subgraphs(graphs[[i]])
  ptrn <- paste0("^", sample_names[i], "_")
  clusters_ <- cluster_ids[grepl(ptrn, names(cluster_ids))]
  names(clusters_) <- gsub(ptrn, "", names(clusters_))

  spatial[[sample_names[i]]] <- tibble(
    sample_by_cell = paste0(sample_names[i], "_", names(SG)),
    entropy = entropies(SG, clusters_),
    nb_size = avg_dists(SG)
  )
}
```

```{r plot_spatial}
spatial <- bind_rows(spatial)

ggplot(spatial %>% inner_join(cd)) +
  geom_histogram(aes(x = entropy, fill = as.factor(TIL_score))) +
  scale_fill_brewer(palette = "Greens", na.value = "grey") +
  facet_grid(cell_group ~ ., scale = "free")

ggplot(spatial %>% inner_join(cd)) +
  geom_histogram(aes(x = nb_size, fill = as.factor(TIL_score))) +
  scale_fill_brewer(palette = "Greens", na.value = "grey") +
  facet_grid(cell_group ~ ., scale = "free")
```


```{r}
spatial_samp <- spatial %>%
  inner_join(cd) %>%
  group_by(SampleID) %>%
  summarise(
    me = mean(entropy),
    sde = sd(entropy),
    mnb = mean(nb_size),
    sdnb = sd(nb_size)
  ) %>%
  left_join(cd_samp)

ggplot(spatial_samp) +
  geom_point(
    aes(x = me, y = Survival_days_capped_2016.1.1, size = TIL_score)
  )
```


