---
title: Cell-Level Analysis
BiocStyle::html_document
---

```{r}
library("dplyr")
library("forcats")
library("ggplot2")
library("readr")
library("reshape2")
library("tibble")
library("tidyr")
theme_set(theme_bw() + theme(panel.grid=element_blank()))
```

# (Joining and ) Subsampling [Mass Spec]

The data are given to us split by cell types, because they each have a different
set of markers (apparently[^1 there are a few annoying discrepancies, like
`HR_DR` in `tcell.sce` and `HLADR` in `livecells.sce`]). We'll join them
together, and put NA for whenever a particular channels is not available.

This joins the row and column data, but not the assays.
```{r eval=FALSE}
data_dir <- file.path("..", "Data")
load(file.path(data_dir, "masstagSCE.rda"))
cell_types <- ls()[grep("sce", ls())]
sapply(cell_types, function(x) get(x))

row_union <- lapply(cell_types, function(x) rowData(get(x)))
row_union <- do.call(rbind, row_union) %>%
  as_tibble() %>%
  unique() %>%
  arrange(channel_name, marker_name)

col_union <- lapply(cell_types, function(x) as_tibble(colData(get(x))))
names(col_union) <- cell_types
col_union <- bind_rows(col_union, .id="cell_type") %>%
  select(sample_id, starts_with("patient"), everything()) %>%
  select(-file_name) %>%
  mutate_at(vars(matches("Age|percent|Score")), as.numeric) %>%
  mutate_at(vars(-matches("Age|percent|Score")), as_factor)
```

```{r eval=FALSE}
glimpse(row_union)
summary(col_union)
```

Now we'll join the assays across cell types.

```{r eval = FALSE}
library("Matrix")
X <- list()

for (i in seq_along(cell_types)) {
  print(cell_types[i])
  X[[i]] <- get(cell_types[i]) %>%
    assay() %>%
    t() %>%
    as_tibble()
  rm(list = cell_types[i])
}

X <- bind_rows(X) %>%
  as.matrix() %>%
  Matrix(sparse=TRUE)

sample_ix <- sample(seq_len(nrow(X)), 1e4)
X <- X[sample_ix, ]
X[is.na(X)] <- 0
col_union <- col_union[sample_ix, ]
```

```{r eval = FALSE}
saveRDS(X, file.path(data_dir, "mass_spec_sample.rds"))
write_csv(col_union, file.path(data_dir, "../Data/cell_info_sample.csv"))
write_csv(row_union, file.path(data_dir, "../Data/marker_info_sample.csv"))
```

# Cell-Level [Mass Spec]

Reading in the subsampled data...
```{r}
X <- readRDS(file.path(data_dir, "mass_spec_sample.RDS"))
row_union = read_csv(file.path(data_dir, "marker_info_sample.csv"))
col_union <- read_csv(file.path(data_dir, "cell_info_sample.csv"))
```

We can make standard pair plots and a heatmap. The heatmap is a bit artificial, since we joined across cell types which didn't all have the same markers.
```{r}
clust_cells <- hclust(dist(X))
clust_markers <- hclust(dist(t(X)))

plot_ix <- c(1, 9, 17, 25, 33, 41, 49, 57, 65, 74)
for (i in plot_ix) {
  pairs(
    as.matrix(X[, clust_markers$order[plot_ix[i]:plot_ix[i + 1]]]),
    cex=0.1, col=rgb(0, 0, 0, 0.1)
  )
}

heatmap(as.matrix(X))
```


Let's refresh ourselves about the phenotypic data, before trying to link it up
with the assays.
```{r}
ggplot(col_union) +
  geom_histogram(aes(x = Health.Status), stat="count")
ggplot(col_union) +
  geom_histogram(aes(x = Histopathology), stat="count") +
  theme(axis.text.x = element_text(angle = -90))
ggplot(col_union) +
  geom_histogram(aes(x = Clinical.Subtype), stat="count")
ggplot(col_union) +
  geom_histogram(aes(x = ER.IRS.Score))
ggplot(col_union) +
  geom_histogram(aes(x = ER.percent.positive.cells.by.IHC))
ggplot(col_union) +
  geom_violin(aes(x = Clinical.Subtype, y = ER.IRS.Score))

u <- col_union[, sapply(col_union, is.numeric)]
for (j in seq_len(ncol(u))) {
  iqr <- IQR(u[[j]], na.rm=T)
  u[, j] <- u[, j] + runif(nrow(u), -0.15 * iqr, 0.15 * iqr)
}

pairs(u, cex=0.1, col=rgb(0, 0, 0, 0.1))
rm(u, iqr)

ggplot(col_union) +
  geom_jitter(
    aes(
      x = Histopathology,
      y = PR.IRS.Score,
      col = Clinical.Subtype
    ),
    size = 0.1,
    alpha = 0.2
  ) +
  scale_x_discrete(label = function(x) abbreviate(x, minlength=10)) +
  facet_grid(~Clinical.Subtype, scales="free_x") +
  theme(axis.text.x = element_text(angle = -90)) +
  theme(legend.position = "none")
```

Okay, now let's join it back in with the assays.
