---
title: "Exploratory Plots"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  BiocStyle::html_document
---

```{r}
library("ggplot2")
library("dplyr")
library("reshape2")
library("tibble")
library("forcats")
theme_set(theme_bw() + theme(panel.grid=element_blank()))
```

Let's make some plots.

Loading the data.

```{r}
data_dir <- file.path("..", "Data")
load(file.path(data_dir, "masstagSCE.rda"))
load(file.path(data_dir, "mibiSCE.rda"))
```

# CD45 Data

```{r}
dim(cd45.sce) # CD45's by cells
dim(rowData(cd45.sce)) # CD45 info
dim(colData(cd45.sce)) # cell info
```

Looks like each channel corresponds to exactly one marker name, and one hgnc
symbol. So, the `rowData` is really just a dictionary lookup, it is not data.

```{r}
row_data <- rowData(cd45.sce) %>%
  as.data.frame() %>%
  as_tibble()

summary(row_data)
```

I'm not going to replace the `na`'s in the `colData` because I think they might be informative.
```{r}

fct_cols <- which(sapply(colData(cd45.sce), is.factor))
col_data <- colData(cd45.sce) %>%
  as.data.frame() %>%
  as_tibble() %>%
  mutate_all(as.factor) %>%
  mutate_at(vars(matches("Score"), matches("percent"), matches("Age")), as.numeric)

summary(col_data)

for (j in seq_len(ncol(col_data))) {
  cur_col = colnames(col_data)[j]
  p <- ggplot(col_data) +
    geom_histogram(
      aes_string(
        x = cur_col
      ),
      stat = "count"
    ) +
    ggtitle(cur_col) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 0),
      )

  if (is.factor(col_data[[j]])) {
    p <- p +
      scale_x_discrete(label = function(x) abbreviate(x, minlength=10))
  }

  print(p)
}

```

Let's look at the assays, at least the marginals.

```{r fig.hold="hold", out.width="15%"}
x <- assay(cd45.sce)
dim(x)

for (j in seq_len(nrow(x))) {
  hist(x[j, ], breaks=100, main=rownames(x)[j])
}
```

There are all sorts of ways to combine the rowdata with the assay, to see
interesting stuff. Here are some examples.

```{r}
random_patients <- sample(levels(merged_cx$patient_id.x), 10)
s_ix <- which(col_data$patient_id.x %in% random_patients)
x_small <- x[, s_ix]
c_small <- col_data[s_ix, ]

merged_cx <- cbind(t(x_small), c_small) %>%
  melt(measure.vars = rownames(x), variable.name = "channel")
```

```{r}
p <- ggplot(merged_cx) +
  geom_violin(
    aes(
      x = Clinical.Subtype,
      y = value,
      fill = Grade
    ),
    alpha=0.3,
  ) +
  facet_wrap(~ channel)

p + scale_y_sqrt()


ggplot(merged_cx) +
  geom_point(
    aes(
      x = Ki.67.percent.positive.cells.by.IHC,
      y = value,
      col = Grade
    ),
    position = position_jitter(w = 0.5),
    size= .5,
    alpha=0.05,
    ) +
  facet_wrap(~ channel) +
  scale_y_sqrt()
```

##

Now we can look at the MIBI data.

```{r}
y <- assay(mibi.sce)

sce_mibi <- as_data_frame(t(x)) %>%
  mutate(source = "sce") %>%
  rownames_to_column("cell_id") %>%
  melt(id.vars = c("cell_id", "source"), variable.name = "channel") %>%
  rbind(
    as_data_frame(t(y)) %>%
    mutate(source = "mibi") %>%
    rownames_to_column("cell_id") %>%
    melt(id.vars = c("cell_id", "source"), variable.name = "channel")
  )

cds <- intersect(rownames(x), rownames(y))
ggplot(sce_mibi %>% filter(channel %in% cds, value > 0)) +
  geom_histogram(
    aes(
      x = sqrt(value),
      fill = source
    ),
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_grid(source~channel, scale="free") +
  ggtitle("Dstn of shared CD's, positive only")

ggplot(sce_mibi %>% filter(channel %in% cds)) +
  geom_histogram(
    aes(
      x = value <= 0, # many mibi values are < 0
      fill = source
    ),
    stat="count"
    ) +
  facet_grid(source~channel, scale="free") +
  ggtitle("Dstn of shared CD's, presence / absence")
```

of course we should also try matching histograms on matching subsets (only grade
3 cancers with other grade 3 cancer cells etc.)

to be continued....
