---
title: Features from MIBI Segmentations
---

For the MIBI data, we have segmentation masks telling us where the cells are
located. The masks themselves are not interesting. However, they give
information about cell-ecosystem heterogeneity and composition which *is* very
interesting. This script produces some intermediate outputs that get more
directly at those questions,

* Split-Channel Segmentations: Each channel in this array is a segmentation mask
  associated with a particular cell type, derived from the `cellData` table. It
  makes it easier to compute the density of different types.
* Shared-Border Graphs: For each cell, we can identify the cells with which it
  shares a border. Each node is associated with a cell type and ID. Ideally, the
  edges would be weighted by the length of the shared border, though I'm not
  sure if that's easy to compute.
* Moving Window Graphs: The shared-border approach will ignore cells that are
  near, but not bordering, others. This approach instead links two cells if they
  are within a distance $\eps$ of one another.
* Moving Window Statistics: For each window, we can different cell densities.

```{r}
library("ggplot2")
library("dplyr")
library("reshape2")
library("tibble")
library("readr")
library("forcats")
library("raster")
theme_set(theme_bw() + theme(panel.grid=element_blank()))
```

```{r}
data_dir <- file.path("..", "Data", "TNBC_shareCellData")
f <- list.files(data_dir, "*tif", full.names = TRUE)[1]
im <- raster(f)
## im <- crop(im, extent(im, 0, 250, 0, 250))
```

Let's convert the image into polygons

```{r}
library("stars")
polys <- st_as_stars(im) %>%
  st_as_sf(merge = TRUE) %>%
  st_cast("POLYGON")
```

Now let's join polygons that are neighbors. Notice that there is a thin border
between cells. We'll need to ignore that when defining neighbors.

```{r fig.width=14, fig.height=14}
library("spdep")
cell_data <- read_csv(file.path(data_dir, "cellData.csv")) %>%
  filter(SampleID == 1)

polys0 <- polys %>%
  inner_join(cell_data, by = c("p1_labeledcellData" = "cellLabelInImage")) %>%
  group_by(p1_labeledcellData) %>% # some polygons are individual pixels at corners --> take union
  summarise_all(function(x) {x[1]})

nb_contig <- poly2nb(polys0, snap=1)
coords <- st_coordinates(st_centroid(st_geometry(polys0)))

plot(polys0$geometry[1:nrow(polys0)])
plot(nb_contig, coords, add=T)
```

Alternatively, we can compute the k nearest neighbors for each point, even if
they aren't bordering any cells themselves.

```{r fig.width=14, fig.height=14}
nb_knn <- knn2nb(knearneigh(coords, k = 5))
plot(polys0$geometry[1:nrow(polys0)])
plot(nb_knn, coords, add=T)
```

Now, let's link up the cell identities with these graphs -- we want to see how
mixed in the different cell types are with one another, for example.

```{r}
ggplot(polys0) +
  geom_sf(aes(fill=as.factor(immuneGroup))) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))
```

We can build a graph that summarizes this information.

```{r}
library("igraph")
relations_data <- list()
for (i in seq_along(nb_contig)) {
  relations_data[[i]] <- tibble(
    from = polys0$p1_labeledcellData[i],
    to = polys0$p1_labeledcellData[nb_contig[[i]]]
  )
}
relations_data <- bind_rows(relations_data)

G <- graph_from_data_frame(relations_data, vertices=polys0)

plot(
  G,
  vertex.color = as.factor(V(G)$immuneGroup),
  vertex.size=3,
  edge.arrow.size=0,
  vertex.label=NA
)
```
