---
title: "Integrating proteomics data"
author: "Pratheepa Jeganathan, Susan Holmes"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  BiocStyle::html_document
params:
  K: 3
  R: 100
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 20,
  fig.height = 15)
```

```{r read_arg}
K <- params$K
K
iter <- params$R
iter
```

```{r}
library(SingleCellExperiment)
library(ggplot2)
library(rstan)
library(plyr)
library(reshape2)
library(readr)
library(magrittr)
library(MultiAssayExperiment)
library(dplyr)
library(DESeq2)
library(abind)
library(tibble)
library(RColorBrewer)
library(raster)
library(stringr)
library(ggthemes)
library(pheatmap)
library(gtools)
```

# CyTOF
- Wagner 2018 Cytof
  - 140 breast cancer patients; Of 140, 6 tripple negative (TN)
  - cd45_sce_dropna$Clinical.Subtype == "TN
  - 3 cancer-free
  - 73 protein markers in immune-centric and tumor-centric microenvironment
  
- Keren 2018 Multiplex Ion Bean Imaging (MIBI)
  - Tumor-immune microenvironment in TN patients
  - 41 TN patients
  - 36 proteins
  
We will choose patients with TN with immune and tumor cells in both CyTOF and MIBI for the integrative analysis.

We will choose one patient from CyTOF (BB028) and one patient from MIBI. 

We make a multi-assay experiemnt object.

We do inverse transformation of cytof data.

We use the upper limit of cytof data to scale the mibi (normalized) data to the range of cytof data. 

After making same units in both cytof and mibi, we impute the protein expression that is not measured in one of the platform.

We round the protein expressio to integer.

We use topic modeling to find the dominated topics in cytof and mibi cells.

We assign the spatial location of mibi cells to the cytof data if the difference between the topic distribution of the dominated topic in the cytof and the dominated topic in the mibi is the minimum.

We infer spatial location of cytof cells using the mibi (one patient, but we can increase the number of patients).

We plot the spatial co-location of cytof cells related to the mibi spatial location.


## Join the assays and impute
```{r}
mae <- readRDS("../Results/mae_mibi_cytof_one.rds")
x1 <- assay(mae[["cd45"]]) %>% t() %>% data.frame()
x2 <- assay(mae[["mibi"]]) %>% t() %>%  data.frame()
x <-  full_join(x1, x2)
markers <- colnames(x)
cells <- c(rownames(x1), rownames(x2))
common_markers <- colnames(x1)[colnames(x1) %in% colnames(x2)]
common_markers
#rm(x1,x2) 

# some features are not recorded in one domain - impute data after scaling
colMissing <- apply(x, 2, function(y){mean(is.na(y))})

temp <- x %>% as.matrix() %>% t()
rm(x1, x2)
```

```{r eval=FALSE}
imdata <- impute::impute.knn(temp, k=5, rowmax = 0.5, colmax = 0.9)$data # temp is variables in the rows and samples in the columns
# x[is.na(x)] <- 0
saveRDS(imdata, "../Results/imdata_mibi_cytof_one.rds")

```

```{r}
imdata <- readRDS("../Results/imdata_mibi_cytof_one.rds")
x <- imdata %>% t()
rownames(x) <- cells
rm(imdata, temp)
```


# Topic modeling


Consider test cells
```{r eval=FALSE}
test_samples <- sample(rownames(x2), size = round(length(rownames(x2))*0.1, digits = 0))
saveRDS(test_samples, file = "../Results/test_samples_one_mibi_one_cytof.rds")
```

```{r}
test_samples <- readRDS("../Results/test_samples_one_mibi_one_cytof.rds")
x_test_train <- x
x_test <- x[test_samples, ]
x_test <- apply(x_test , c(1,2), function(y){as.integer(y)})
x_test <- as.matrix(x_test)
mae_test <- mae[, test_samples,]

x_train <- x[!(rownames(x) %in% test_samples), ]
x_train <- apply(x_train , c(1,2), function(y){as.integer(y)})
x_train <- as.matrix(x_train)
mae_train <- mae[, which(!(rownames(x) %in% test_samples)),]
```


```{r eval=FALSE}
# theta[d] ~ dirichlet(alpha), alpha pseudocount for each topic
# beta[k] ~ dirichlet(gamma), gamma pseudocount for each ASV in each topic
stan.data <- list(K = K, 
  V = ncol(x_train), 
  D = nrow(x_train), 
  n = x_train, 
  alpha = rep(.8, K), 
  gamma = rep(.5, ncol(x_train))
)
```

```{r}
fileN <- paste0("../Results/LDA_mibi_cytof_one_K_",K,"_ite_",iter,".RData")
fileN
```



```{r eval=FALSE}
t1 <- proc.time()
stan.fit <- stan(file = "./lda.stan", 
  data = stan.data, 
  iter = iter, 
  chains = 4, 
  sample_file = NULL,
  diagnostic_file = NULL,
  cores = 4,
  control = list(adapt_delta = 0.9),
  save_dso = TRUE,
  algorithm = "NUTS")
proc.time() - t1

save(stan.fit, file = fileN)
```

```{r}
load(file = fileN)
```



Sampler diagnostics
```{r}
sampler_params <- get_sampler_params(stan.fit, 
                                     inc_warmup = FALSE)
colnames(sampler_params[[1]])

mean_accept_stat_by_chain <- sapply(sampler_params, 
                                    function(x) mean(x[, "accept_stat__"]))
mean_accept_stat_by_chain

max_treedepth_by_chain <- sapply(sampler_params, 
                                 function(x) max(x[, "treedepth__"]))
max_treedepth_by_chain
```


# Visualization 

## Extract posterior samples
```{r}
samples <- rstan::extract(stan.fit, 
                          permuted = TRUE, 
                          inc_warmup = FALSE, 
                          include = TRUE)# samples is a list
rm(stan.fit)
```

## Alignment
- Create a Topic $*$ Chain matrix
```{r}
source("../R_BANF/alignmentMatrixMAE.R")
source("../R_BANF/thetaAligned.R")
theta <- samples$theta 
# theta <- theta[c(1:25, 51:75, 101:125, 151:175) , , ]

# aligned <- alignmentMatrixMAE(theta, 
#                            mae_train, 
#                            K, 
#                            iter = iter, 
#                            chain = 4,
#                            SampleID_name = "cell_id")
# theta_aligned <- thetaAligned(theta, 
#                               K, 
#                               aligned, 
#                               iter = iter, 
#                               chain = 4)
theta_aligned <- theta[1:(iter/2), ,]
dimnames(theta_aligned)[[2]] <- mae_train$cell_id
dimnames(theta_aligned)[[3]] <- c(paste0("Topic_", seq(1,K)))

# array to a dataframe
theta_all <- melt(theta_aligned)
colnames(theta_all) <- c("iteration", "Sample", "Topic", "topic.dis")
theta_all$Chain <- paste0("Chain ", rep(seq(1, 1), each = (iter/2)))

sam <- colData(mae_train) %>% data.frame()
theta_all$Sample <- as.character(theta_all$Sample)
theta_all <- left_join(theta_all, sam, by =c("Sample"= "cell_id"))
theta_all$Chain <- factor(theta_all$Chain)
theta_all$Topic <- factor(theta_all$Topic)
theta_all$Sample <- factor(theta_all$Sample)

theta_all$cell_type <- factor(theta_all$cell_type)
theta_all$method <- ifelse(is.na(theta_all$cell_type), "cytof", as.character(theta_all$cell_type))
```



```{r}
manual_col <- tableau_color_pal("Classic 20")(length(unique(theta_all$method)))

theta_summary <- theta_all %>% 
  group_by(Sample, Topic, method) %>% 
  summarize(median.topic.dis = median(topic.dis)) %>% 
  ungroup() %>% 
  mutate(Topic = factor(Topic, levels = rev(str_c("Topic_",1:K))))

#theta_summary <- dplyr::filter(theta_summary, cell_type %in% c("Macrophages", "Tumor"))

sample_cells <- unique(theta_summary$Sample)
sample_cells_select <- sample(sample_cells, 50)

theta_summary <- dplyr::filter(theta_summary, 
                               Sample %in% sample_cells_select)

p <- ggplot(theta_summary, 
           aes(x = method, 
               y = Topic, 
               fill = method))
p <- p+
  geom_tile(aes(alpha = median.topic.dis))+
  facet_grid(.~Sample, scale = "free")+
  xlab("method") +
  scale_fill_manual(name = "method", 
                    values = manual_col) +
  scale_alpha(name = "median topic distribution") +
  theme_minimal(base_size = 20) +  
  theme(plot.title = element_text(hjust = 0.5), 
        strip.text.x = element_text(angle = 90), 
        axis.text.x=element_blank()) 
p

ggsave(paste0("../Figures/topic_dis_mibi_cytof_one_K_",K, ".png"), p, width = 20, height = 10)
rm( p)
```



## Marker distribution
```{r}
source("../R_BANF/betaAligned.R")
beta <- samples$beta # an array (iterations *topic * marker)
# beta_aligned <- betaAligned(beta, 
#                             K, 
#                             aligned, 
#                             iter = iter, 
#                             chain = 4) # an array (iterations *topic * ASV)
beta_aligned <- beta[1:(iter/2), , ]
# array to data frame
beta_hat <- beta_aligned %>%
  melt(varnames = c("iterations", "topic", "marker_ix"), 
       value.name = "beta_h") %>% as_tibble()
beta_hat$marker <- colnames(x_train)[beta_hat$marker_ix]


# join rowData with beta_hat
marker_info <- full_join(rowData(mae_train[["cd45"]]) %>% data.frame(), rowData(mae_train[["mibi"]]) %>% data.frame())
marker_info$marker <- marker_info$marker_name


beta_hat <- beta_hat %>%
  left_join(marker_info, by = "marker") %>% mutate(topic = paste("Topic", topic))


beta_hat$marker <- factor(beta_hat$marker)
beta_hat$marker_ix <- factor(beta_hat$marker_ix)
beta_hat$topic <- factor(beta_hat$topic)


beta_summary <- beta_hat %>% 
  dplyr::group_by(marker_ix, topic) %>% 
  dplyr::summarise(
    marker = marker[1],
    beta_median = median(beta_h),
    marker = marker[1],
    hgnc_symbol = hgnc_symbol[1]
  )

beta_subset <- beta_summary
beta_subset$marker_ix <- rep(seq_len(nrow(beta_subset) / K), each = K)
rm(beta_summary, beta_hat)
```


```{r}
beta_subset <- beta_subset %>% 
  arrange(marker_ix, topic)

beta_subset <- beta_subset %>%
  mutate(Class = factor(marker, levels = unique(beta_subset$marker)),
         Topic = str_remove(topic, "Topic "))


beta_subset$Topic <- factor(beta_subset$Topic, levels = seq(1,K) %>% as.character())

p <- ggplot(beta_subset, 
            aes(x = Topic, 
                y = marker, 
                fill = beta_median)) +
  geom_tile() +
  ylab("Marker")+
  scale_fill_gradientn(name = "Median marker distribution", 
                       colours = c("gray98", "dodgerblue")) +  
  theme_minimal(base_size = 20) +
  theme(plot.title = element_text(hjust = 0.5)) 

p
ggsave(paste0("../Figures/fea_dist_barplot_mibi_cytof_one_K_",K, ".png"), p, width = 20, height = 11)
rm(p)
```



# Computing statistics after alignment

## Compute predictive log likelihood in training set

$$\left. K_{.j} \right \vert S_{j}, \theta_{j} , B  \sim  \text{Multinomial}\left( S_{j} , B  \theta_{j} \right), j =1, \cdots, N,$$
```{r}
log_lik_total_for_each_iteration <- numeric()
for(it in 1:((iter/2)*1)){
  log_lik <- numeric()
    for(j in 1:dim(x_train)[1]){# For each sample j in the first posterior sample
      p_vec_pos <- as.matrix(t(beta_aligned[it, , ])) %*% matrix(theta_aligned[it, j,], nrow = K, byrow = T)
      
      log_lik[j] <- dmultinom(x_train[j, ], size = sum(x_train[j, ]), prob = p_vec_pos, log = TRUE)
    }
  log_lik_total_for_each_iteration[it] <- sum(log_lik)
}


df_lp_corrected <- data.frame(lp = log_lik_total_for_each_iteration, 
                             Chain = paste0("Chain ", rep(seq_len(1), each = (iter/2))))


fileN <- paste0("../Results/df_lp_train_mibi_cytof_one_K_", K, ".rds")
saveRDS(df_lp_corrected, fileN)

p_lp <- ggplot(data = df_lp_corrected) +
  geom_histogram(aes(x = lp, col = Chain, fill = Chain), 
                 bins = 100) + 
  theme_minimal(base_size = 20)
p_lp

ggsave(paste0("../Figures/lp_train_mibi_cytof_one_K_", K, ".png"), p_lp, width = 9, height = 6)
rm(log_lik_total_for_each_iteration, df_lp_corrected, p_lp)
```


```{r eval=FALSE}
df_lp_corrected_all_K <- data.frame()
for(top in c(3:16, 20)){
  fileN <- paste0("../Results/df_lp_train_mibi_cytof_one_K_", top, ".rds")
  df_lp_corrected <- readRDS(fileN)
  df_lp_corrected$Num_Topic <- paste0("T=", rep(top, (iter/2*1)))
  df_lp_corrected_all_K <- rbind(df_lp_corrected_all_K, df_lp_corrected)
}

df_lp_corrected_all_K$Num_Topic <- factor(df_lp_corrected_all_K$Num_Topic, levels = c(paste0("T=", c(seq(3, 16), 20))
))

p <- ggplot(data = df_lp_corrected_all_K) + 
  geom_boxplot(aes(x = Num_Topic, y = lp)) + 
  ylab("Predictive log-likelihood in train data") + 
  xlab("Number of topics")

p
ggsave("../Figures/lp_train_mibi_cytof_one.png", p, width = 7, height = 4)
```


## Compute $\hat{R}$

```{r eval=FALSE}
Rhat_theta <- matrix(nrow = dim(theta_aligned)[2], ncol = dim(theta_aligned)[3])
ESS_bulk_theta <- matrix(nrow = dim(theta_aligned)[2], ncol = dim(theta_aligned)[3])
ESS_tail_theta <- matrix(nrow = dim(theta_aligned)[2], ncol = dim(theta_aligned)[3])
for(sam in 1:dim(theta_aligned)[2]){
  for(top in 1:dim(theta_aligned)[3]){
    sims_theta <- matrix(theta_aligned[ ,sam , top], nrow = (iter/2), ncol = 4, byrow = FALSE)
    Rhat_theta[sam, top] <- Rhat(sims_theta)
    ESS_bulk_theta[sam, top] <- ess_bulk(sims_theta)
    ESS_tail_theta[sam, top] <- ess_tail(sims_theta)
  }
  
}

Rhat_theta <- as.vector(Rhat_theta)
ESS_bulk_theta <- as.vector(ESS_bulk_theta)
ESS_tail_theta <- as.vector(ESS_tail_theta)

Rhat_beta <- matrix(nrow = dim(beta_aligned)[2], ncol = dim(beta_aligned)[3])
ESS_bulk_beta <- matrix(nrow = dim(beta_aligned)[2], ncol = dim(beta_aligned)[3])
ESS_tail_beta <- matrix(nrow = dim(beta_aligned)[2], ncol = dim(beta_aligned)[3])
  
for(top in 1:dim(beta_aligned)[2]){
    for(fea in 1:dim(beta_aligned)[3]){
      sims_beta <- matrix(beta_aligned[ , top, fea], nrow = (iter/2), ncol = 4, byrow = FALSE)
      Rhat_beta[top, fea] <- Rhat(sims_beta)
      ESS_bulk_beta[top, fea] <- ess_bulk(sims_beta)
      ESS_tail_beta[top, fea] <- ess_tail(sims_beta)
    }
  
}

Rhat_beta <- as.vector(Rhat_beta)
ESS_bulk_beta <- as.vector(ESS_bulk_beta)
ESS_tail_beta <- as.vector(ESS_tail_beta)

Rhat <- c(Rhat_theta, Rhat_beta)
fileN <- paste0("../Results/Rhat_mibi_cytof_one_K_",K, ".rds")
saveRDS(Rhat, fileN)

ESS_bulk <- c(ESS_bulk_theta, ESS_bulk_beta)
fileN <- paste0("../Results/ESS_bulk_mibi_cytof_one_K_",K, ".rds")
saveRDS(ESS_bulk, fileN)

ESS_tail <- c(ESS_tail_theta, ESS_tail_beta)
fileN <- paste0("../Results/ESS_tail_mibi_cytof_one_K_",K, ".rds")
saveRDS(ESS_tail, fileN)

# R hat ~ 1.05
p_rhat <- ggplot(data.frame(Rhat = Rhat)) +
  geom_histogram(aes(x = Rhat), 
                 fill = "lavender", 
                 colour = "black", 
                 bins = 100) +
  theme_minimal(base_size = 20) +  
  theme(plot.title = element_text(hjust = 0.5))  +
  theme_minimal(base_size = 20) +
  xlab("")
p_rhat

ggsave(paste0("../Figures/Rhat_mibi_cytof_one_K_",K, ".png"), p_rhat, width = 9, height = 6)

rm(Rhat_theta, ESS_bulk_theta, ESS_tail_theta, Rhat_beta, ESS_bulk_beta, ESS_tail_beta, p_rhat,Rhat)
```

## Compute ESS

```{r eval=FALSE}
# ESS bulk and ESS tail at least 100 per Markov Chain in order to be reliable and indicate that estimates of respective posterior quantiles are reliable

p_ess_bulk <- ggplot(data.frame(ESS_bulk = ESS_bulk)) +
  geom_histogram(aes(x = ESS_bulk), 
                 fill = "lavender", 
                 colour = "black", 
                 bins = 100) +
  theme_minimal(base_size = 20) +  
  theme(plot.title = element_text(hjust = 0.5))   + 
  theme_minimal(base_size = 20) +
  xlab("")

# + 
#   xlab("Effective sample size (ESS)")
p_ess_bulk

ggsave(paste0("../Figures/ess_bulk_mibi_cytof_one_K_",K, ".png"), p_ess_bulk, width = 9, height = 6)

p_ess_tail <- ggplot(data.frame(ESS_tail = ESS_tail)) +
  geom_histogram(aes(x = ESS_tail), 
                 fill = "lavender", 
                 colour = "black", 
                 bins = 100) +
  theme_minimal(base_size = 20) +  
  theme(plot.title = element_text(hjust = 0.5))

p_ess_tail
ggsave(paste0("../Figures/ess_tail_mibi_cytof_one_K_",K, ".png"), p_ess_tail, width = 9, height = 6)

rm(ESS_bulk, p_ess_bulk, ESS_tail, p_ess_tail)
```

# Model assesment on simulated data
```{r eval=FALSE}
# Simulated data
x_sim <- samples$x_sim # iteration * samples * ASVs
# Choose only the first chain
x_sim <- x_sim[1:(iter/2), ,] # For each iteration, simulated data is x_sim[i, ,]
x_max_asv <- apply(x %>% data.frame , 2, median)
sim_ite <- dim(x_sim)[1]
ite_plot <- 19# or ite_plot <- sim_ite
#Find maximum of each asv for each iteration
max_all <- data.frame(x_max_asv = x_max_asv)
for(i in 1:ite_plot){
  x_sim_i <- x_sim[i, ,]
  max_x_sim_i <- apply(x_sim_i %>% data.frame , 2, median)
  max_all <- cbind(max_all, max_x_sim_i)
}

colnames(max_all) <- c("x_med_obs", paste0("x_med_sim_1", seq(1, ite_plot)))
avg_max_square <- numeric()

# histogram of maximum
max_all_long <- melt(max_all)
max_all_long$type <- c(rep("observed", dim(x_sim)[3]), rep("simulated", dim(x_sim)[3]*(ite_plot)))
p_hist_obs_sim <- ggplot(data = max_all_long) + 
  geom_histogram(aes(x = value, group = variable, fill = type), 
                 colour = "black",
                 bins = 100) + 
  facet_wrap(~variable, nrow = 5) +
  theme(legend.position = "none")

p_hist_obs_sim
ggsave(paste0("../Figures/model_assesment_hist_obs_sim_mibi_cytof_one_K_",K, ".png"), p_hist_obs_sim, width = 12, height = 6)

rm(x_sim, samples, p_hist_obs_sim)
```

The distribution of observed maximum of each protein expression (green color) is similar to maximum of simulated data from the fitted topic model.


# Log posterior on the test data

```{r eval=FALSE}
#x_test
dimnames(x_test) = NULL
test_data = list(K = K,
  V = ncol(x_test),
  D = nrow(x_test),
  n = x_test,
  alpha = rep(.8, K),
  gamma = rep(.5, ncol(x_test))
)

log_lik_total_for_each_iteration <- numeric()
for(it in 1:((iter/2)*1)){
  log_lik <- numeric()
    for(j in 1:dim(x_test)[1]){# For each sample j in the test set
      theta_aligned_sim <- rdirichlet(1, test_data$alpha)# simulate topic distribution from prior distribution
      p_vec_pos <- as.matrix(t(beta_aligned[it, , ])) %*% matrix(theta_aligned_sim, nrow = K, byrow = T)
      
      log_lik[j] <- dmultinom(x_test[j, ], size = sum(x_test[j, ]), prob = p_vec_pos, log = TRUE)
    }
  log_lik_total_for_each_iteration[it] <- sum(log_lik)
}


df_lp_corrected <- data.frame(lp = log_lik_total_for_each_iteration, 
                             Chain = paste0("Chain ", rep(seq_len(1), each = (iter/2))))

fileN <- paste0("../Results/df_lp_test_mibi_cytof_one_K_", K, ".rds")
saveRDS(df_lp_corrected, fileN)
rm(x_test, log_lik_total_for_each_iteration, df_lp_corrected)
```


```{r eval=FALSE}
df_lp_corrected_all_K <- data.frame()
for(top in c(3:5, 15, 20)){
  fileN <- paste0("../Results/df_lp_test_mibi_cytof_one_K_", top, ".rds")
  df_lp_corrected <- readRDS(fileN)
  df_lp_corrected$Num_Topic <- paste0("T=", rep(top, (iter/2*1)))
  df_lp_corrected_all_K <- rbind(df_lp_corrected_all_K, df_lp_corrected)
}

df_lp_corrected_all_K$Num_Topic <- factor(df_lp_corrected_all_K$Num_Topic, levels = c(paste0("T=", c(seq(3, 5), 15, 20))
))

p <- ggplot(data = df_lp_corrected_all_K) + 
  geom_boxplot(aes(x = Num_Topic, y = lp)) + 
  ylab("Predictive log-likelihood in test data") + 
  xlab("Number of topics")

p
ggsave("../Figures/lp_test_mibi_cytof_one.png", p, width = 7, height = 4)
```

# Infer spatial location of cytof immune cells

based on the median topic distribution identified the most dominated topic in each cell
```{r}
theta_summary <- theta_all %>% 
  group_by(Sample, Topic, method) %>% 
  summarize(median.topic.dis = median(topic.dis)) %>% 
  ungroup() %>% 
  mutate(Topic = factor(Topic, levels = rev(str_c("Topic_",1:K))))

theta_summary_2 <- theta_summary %>% 
  dplyr::group_by(Sample) %>% 
  dplyr::summarise(
    max_topic = max(median.topic.dis),
    Topic = Topic[which(median.topic.dis == max(median.topic.dis))],
    method = method[1]
  )

saveRDS(theta_summary_2, paste0("../Results/co_location_mibi_cytof_one_K_", K, ".rds"))
```

MIBI has only one site so we may not be able to infer spatial location for some cells in CyTOF.

```{r}
# mibi X, Y coordinates
df <- data.frame(Sample = mae_train$cell_id, X = mae_train$centroid_X, Y = mae_train$centroid_Y)
df <- left_join(theta_summary_2, df, by = "Sample") %>% data.frame()
df$Sample <- as.character(df$Sample)
rownames(df) <- df$Sample
for(i in 1:dim(df)[1]){
  if(df$method[i] == "cytof"){
    #subset the df mibi cells with the topic for ith ccytof cell
    df_i <- filter(df, Topic == df$Topic[i] & method != "cytof")
    com_cell_close_to_max_topic_i <- df_i$Sample[which(abs(df$max_topic[i] - df_i$max_topic) == min(abs(df$max_topic[i] - df_i$max_topic)))]
    df$X[i] <- df$X[which(df$Sample == com_cell_close_to_max_topic_i)]
    df$Y[i] <- df$Y[which(df$Sample == com_cell_close_to_max_topic_i)]
  }
}
```

Plot the spatial location of cytof cells from patient BB028
```{r}
manual_col <- tableau_color_pal("Classic 20")(length(unique(df$method)))
df$method <- factor(df$method)
df_cytof <- dplyr::filter(df, method == "cytof")
p_co <- ggplot(data = df) + 
  geom_point(aes(x= X, y = Y, col = method), size = 5) +
  theme_minimal() +
  scale_color_manual(values = manual_col)  +
  theme(aspect.ratio = 1, legend.position = "right") +
  labs(fill = "method") +
  xlim(c(0, max(df$X))) +
  ylim(c(0, max(df$Y))) + 
  geom_text(data = df_cytof, aes(x = X, y= Y, label = Sample), check_overlap = TRUE)

p_co

ggsave(paste0("../Figures/co_location_mibi_cytof_one_K_", K, ".png"), p_co, width = 15, height = 15)
```

# Predict spatial pattern of proteins not measured in the MIBI-TOF
```{r eval=FALSE}
proteins_not_measured_mibi <- rowData(mae[["cd45"]])$marker_name[!(as.character(rowData(mae[["cd45"]])$marker_name) %in% as.character(rowData(mae[["mibi"]])$channel_name))] %>% as.character()

# Simulated data
x_sim <- samples$x_sim # iteration * samples * ASVs
# Choose only the first chain
x_sim <- x_sim[1, ,] # For each iteration, simulated data 
colnames(x_sim) <- colnames(x_train)
rownames(x_train) <- rownames(x_train)
ind_mibi <- which(rownames(x_train) %in% colnames(assay(mae_train[["mibi"]])))
x_sim_mibi <- x_sim[ind_mibi, ]
ind_pro <- which(colnames(x_train) %in% proteins_not_measured_mibi)
ind_pro_present <- which(!(colnames(x_train) %in% proteins_not_measured_mibi))

x_sim_mibi_only_measured <- x_train[ind_mibi, ind_pro_present]
x_sim_mibi_only_measured_t <- asinh(x_sim_mibi_only_measured) 

x_sim_mibi_predict <- x_sim_mibi
x_sim_mibi_predict_t <- asinh(x_sim_mibi_predict)

library(uwot)
# UMAP of measured proteins
mibi_umap <- umap(x_sim_mibi_only_measured_t)
umap_df <- data.frame(UMAP1 = mibi_umap[,1], 
                      UMAP2 = mibi_umap[,2], 
                      cell_id = colData(mae_train)$cell_id[ind_mibi])
mibi_cell_data <- colData(mae_train)[ind_mibi, ] %>% data.frame()
umap_df  <- left_join(umap_df, mibi_cell_data, by = "cell_id")

manual_col <- tableau_color_pal("Classic 20")(length(unique(mae_train$cell_type)))

p1 <- ggplot(data = umap_df, 
       aes(x = UMAP1, y = UMAP2, color = cell_type)) + 
  geom_point(size = 3) +
  theme_bw() +
  scale_color_manual(values = manual_col)  +
  labs(color = "Cell types") + 
  ggtitle("UMAP with measured proteins") +
  theme(aspect.ratio = 1, plot.title = element_text(hjust = 0.5))
p1
# UMAP of measured and predicted protein
# UMAP of measured proteins
mibi_umap <- umap(x_sim_mibi_predict_t)
umap_df <- data.frame(UMAP1 = mibi_umap[,1], 
                      UMAP2 = mibi_umap[,2], 
                      cell_id = colData(mae_train)$cell_id[ind_mibi])
mibi_cell_data <- colData(mae_train)[ind_mibi, ] %>% data.frame()
umap_df  <- left_join(umap_df, mibi_cell_data, by = "cell_id")

manual_col <- tableau_color_pal("Classic 20")(length(unique(mae_train$cell_type)))

p2 <- ggplot(data = umap_df, 
       aes(x = UMAP1, y = UMAP2, color = cell_type)) + 
  geom_point(size = 3) +
  theme_bw() +
  scale_color_manual(values = manual_col) +
  theme(aspect.ratio = 1) +
  labs(color = "Cell types")
p2


```


```{r eval=FALSE}
# Simulated data
x_sim <- samples$x_sim # iteration * samples * ASVs
# Choose only the first chain
x_sim <- x_sim[4, ,] # For each iteration, simulated data 
colnames(x_sim) <- colnames(x_train)
rownames(x_train) <- rownames(x_train)
ind_mibi <- which(rownames(x_train) %in% colnames(assay(mae_train[["mibi"]])))
x_sim_mibi <- x_sim[ind_mibi, ]
ind_pro <- which(colnames(x_train) %in% proteins_not_measured_mibi)
ind_pro_present <- which(!(colnames(x_train) %in% proteins_not_measured_mibi))

x_sim_mibi_only_measured <- x_train[ind_mibi, ind_pro_present]
x_sim_mibi_only_measured_t <- asinh(x_sim_mibi_only_measured) 

x_sim_mibi_predict <- x_sim_mibi
x_sim_mibi_predict_t <- asinh(x_sim_mibi_predict)

library(uwot)
# UMAP of measured proteins
mibi_umap <- umap(x_sim_mibi_only_measured_t)
umap_df <- data.frame(UMAP1 = mibi_umap[,1], 
                      UMAP2 = mibi_umap[,2], 
                      cell_id = colData(mae_train)$cell_id[ind_mibi])
mibi_cell_data <- colData(mae_train)[ind_mibi, ] %>% data.frame()
umap_df  <- left_join(umap_df, mibi_cell_data, by = "cell_id")

manual_col <- tableau_color_pal("Classic 20")(length(unique(mae_train$cell_type)))

p11 <- ggplot(data = umap_df, 
       aes(x = UMAP1, y = UMAP2, color = cell_type)) + 
  geom_point(size = 3) +
  theme_bw() +
  scale_color_manual(values = manual_col) +
  theme(aspect.ratio = 1) +
  labs(color = "Cell types")
p11
# UMAP of measured and predicted protein
# UMAP of measured proteins
mibi_umap <- umap(x_sim_mibi_predict_t)
umap_df <- data.frame(UMAP1 = mibi_umap[,1], 
                      UMAP2 = mibi_umap[,2], 
                      cell_id = colData(mae_train)$cell_id[ind_mibi])
mibi_cell_data <- colData(mae_train)[ind_mibi, ] %>% data.frame()
umap_df  <- left_join(umap_df, mibi_cell_data, by = "cell_id")

manual_col <- tableau_color_pal("Classic 20")(length(unique(mae_train$cell_type)))

p22 <- ggplot(data = umap_df, 
       aes(x = UMAP1, y = UMAP2, color = cell_type)) + 
  geom_point(size = 3) +
  theme_bw() +
  scale_color_manual(values = manual_col) +
  labs(color = "Cell types") + 
  ggtitle("UMAP with measured and predicted proteins") +
  theme(aspect.ratio = 1, plot.title = element_text(hjust = 0.5))
p22


```


```{r eval=FALSE}
# Simulated data
x_sim <- samples$x_sim # iteration * samples * ASVs
# Choose only the first chain
x_sim <- x_sim[2, ,] # For each iteration, simulated data 
colnames(x_sim) <- colnames(x_train)
rownames(x_train) <- rownames(x_train)
ind_mibi <- which(rownames(x_train) %in% colnames(assay(mae_train[["mibi"]])))
x_sim_mibi <- x_sim[ind_mibi, ]
ind_pro <- which(colnames(x_train) %in% proteins_not_measured_mibi)
ind_pro_present <- which(!(colnames(x_train) %in% proteins_not_measured_mibi))

x_sim_mibi_only_measured <- x_train[ind_mibi, ind_pro_present]
x_sim_mibi_only_measured_t <- asinh(x_sim_mibi_only_measured) 

x_sim_mibi_predict <- x_sim_mibi
x_sim_mibi_predict_t <- asinh(x_sim_mibi_predict)

library(uwot)
# UMAP of measured proteins
mibi_umap <- umap(x_sim_mibi_only_measured_t)
umap_df <- data.frame(UMAP1 = mibi_umap[,1], 
                      UMAP2 = mibi_umap[,2], 
                      cell_id = colData(mae_train)$cell_id[ind_mibi])
mibi_cell_data <- colData(mae_train)[ind_mibi, ] %>% data.frame()
umap_df  <- left_join(umap_df, mibi_cell_data, by = "cell_id")

manual_col <- tableau_color_pal("Classic 20")(length(unique(mae_train$cell_type)))

p13 <- ggplot(data = umap_df, 
       aes(x = UMAP1, y = UMAP2, color = cell_type)) + 
  geom_point(size = 3) +
  theme_bw() +
  scale_color_manual(values = manual_col) +
  theme(aspect.ratio = 1) +
  labs(color = "Cell types")
p13
# UMAP of measured and predicted protein
# UMAP of measured proteins
mibi_umap <- umap(x_sim_mibi_predict_t)
umap_df <- data.frame(UMAP1 = mibi_umap[,1], 
                      UMAP2 = mibi_umap[,2], 
                      cell_id = colData(mae_train)$cell_id[ind_mibi])
mibi_cell_data <- colData(mae_train)[ind_mibi, ] %>% data.frame()
umap_df  <- left_join(umap_df, mibi_cell_data, by = "cell_id")

manual_col <- tableau_color_pal("Classic 20")(length(unique(mae_train$cell_type)))

p23 <- ggplot(data = umap_df, 
       aes(x = UMAP1, y = UMAP2, color = cell_type)) + 
  geom_point(size = 3) +
  theme_bw() +
  scale_color_manual(values = manual_col) +
  labs(color = "Cell types")+ 
  ggtitle("UMAP with measured and predicted proteins") +
  theme(aspect.ratio = 1, plot.title = element_text(hjust = 0.5))
p23


```

```{r eval=FALSE}
# library(gtable)
# library(grid)
# g1 <- ggplotGrob(p1)
# g2 <-  ggplotGrob(p22)
# g3 <- ggplotGrob(p23)
# g <- cbind(g1, g2)
# 
# g$widths <- unit.pmax(g1$widths, g2$widths)
# grid.newpage()
# grid.draw(g)

library(gridExtra)
com_p <- grid.arrange(p1, arrangeGrob(p22, p23, ncol=1), ncol=2, widths=c(1,1.2))

ggsave("../Figures/predicted_mibi_protein.png", com_p, width = 20, height = 15)
```
