---
title: Creating and Relating Atlases
---

* Single-cell analysis often looks like map making
* Issue is we don't have common features on our two maps
  - only a few shared landmarks

* To make this analogy concrete, we're going to look at the data
* Making atlases
  - Let's show how you can make these atlases, using clustering
  - Also using U-Map or PCA
  - Future: We should also make these using spatial or graph information. So,
    being close by in on an atlas means that you have similar graph features as
    well.
* Interpreting atlases
  - Is over / under representation in some regions of the map related to good /
    poor outcomes?
* Relating atlases
  - Are there similar interpretations coming from the two different datasets? In
    what ways are the stories discrepant?
  - Can we try to map clusters / PCA's onto each other? At the least, we can
    calculate the distance between centroids, using shared features. A more
    sophisticated / imputation-like idea is to link the atlases in a more
    continuous way, using some sort of learned (inverse) function. But let's not
    worry about that.

## Preparing Data

```{r}
library("ggplot2")
library("dplyr")
library("reshape2")
library("tidyr")
library("tibble")
library("forcats")
source("preprocessing.R")
theme_set(theme_bw() + theme(panel.grid=element_blank()))
data_dir <- file.path("..", "Data")
```

While we're developing the analysis, we'll use subsampled versions of all the
data. This loads the masstag and mibi datasets.

```{r}
load(file.path(data_dir, "masstagSCE.rda"))
masstag <- data_list("sce")
load(file.path(data_dir, "mibiSCE.rda"))

masstag <- subsample_experiments(masstag)
mibi <- subsample_experiments(list(mibi.sce), 0.4)[[1]]

colData(mibi)$cell_type <- colData(mibi) %>%
              as.data.frame() %>%
              select(tumor_group, immune_group) %>%
              mutate(
                cell_type = paste0(tumor_group, immune_group),
                cell_type = gsub("not immune", "", cell_type),
                cell_type = gsub("Immune", "", cell_type),
                ) %>%
              .[["cell_type"]] %>%
              as_factor()
```

This next step is optional -- it transforms the original intensities into
quantiles, to make measures comparable across experiments.

First, just to get intuition, look at some transforms.
```{r}
for (j in seq_len(10)) {
  r <- rank(assay(mibi[j, ])) / ncol(mibi)
  plot(assay(mibi[j, ]), r)
}
```

Now we'll make the transformation. (tk should create an rmarkdown parameter to skip this transformation).
```{r}
mibi <- quantile_transform(mibi)
for (i in seq_along(masstag)) {
  masstag[[i]] <- quantile_transform(masstag[[i]])
}
```

## Clustering

* k-means on the raw proteins, separately for each of the cell types.
 - This means that each cluster will have to be marked by celltype-K, rather
   than just K.
 - Another consequence is that we'll have a small set of U-Maps for each
   patient, one for each cell type

```{r}
heatmap(assay(masstag[[1]]))
heatmap(assay(masstag[[2]]))
```

```{r}
K <- 5
masstag_clusters <- lapply(masstag, function(x) { kmeans(t(assay(x)), centers=K) })

mibi_clusters <- list()
mibi_types <- colData(mibi)$cell_type
for (group in unique(mibi_types)) {
  cur_x <- t(assay(mibi)[, mibi_types == group])
  mibi_clusters[[group]] <- kmeans(cur_x, centers=5)
}

mibi_clusters
```

We can visualize the centroids for one of the clusterings. We should do this for
every cluster across all the cell types.

```{r}
centroids_df <- masstag_clusters[[1]]$centers %>%
  melt(varnames = c("cluster", "channel"))

ggplot(centroids_df) +
  geom_tile(
    aes(x = channel, y = cluster, fill = value)
  ) +
  scale_fill_gradient(low = "white", high = "black") +
  theme(axis.text.x = element_text(angle = 0))

cluster_hist_data <- assay(masstag[[1]]) %>%
  t() %>%
  data.frame() %>%
  mutate(
    cell_id = row_number(),
    cluster = as_factor(masstag_clusters[[1]]$cluster)
  ) %>%
  melt(id.vars = c("cell_id", "cluster"))

ggplot(cluster_hist_data) +
  geom_histogram(aes(x = value, fill = cluster), position = "identity", alpha = 0.7) +
  facet_wrap(. ~ variable, ncol = 5) +
  scale_y_continuous(limits = c(0, 750)) +
  theme(panel.spacing = unit(0, "cm"))
```


## Clusters vs. Phenotypes

```{r}
masstag_pheno <- list()
for (i in seq_along(masstag)) {
  masstag_pheno[[i]] <- colData(masstag[[i]]) %>%
    data.frame() %>%
    select(patient_id.x, Clinical.Subtype) %>%
    mutate(
      cluster = as.factor(masstag_clusters[[i]]$cluster),
      cell_type = gsub("\\.sce", "" , names(masstag))[i]
    )
}

masstag_pheno <- bind_rows(masstag_pheno) %>%
  group_by(patient_id.x, Clinical.Subtype, cluster, cell_type) %>%
  summarise(count = n()) %>%
  group_by(patient_id.x) %>%
  mutate(
    total = sum(count),
    prop = count / total
  ) %>%
  unite(cell_cluster, c("cell_type", "cluster"), remove=F)

ggplot(masstag_pheno) +
  geom_boxplot(
    aes(x = Clinical.Subtype, y = prop, fill = Clinical.Subtype),
    varwidth = TRUE
  ) +
  scale_y_sqrt() +
  facet_grid(cell_type~cluster) +
  theme(panel.spacing = unit(0, "cm"))
```

```{r}
masstag_props <- masstag_pheno %>%
  dcast(patient_id.x + Clinical.Subtype ~ cell_cluster, value.var = "prop", fill=0)

sample_clust <- hclust(dist(masstag_props[, -c(1, 2)]))
sample_order <- masstag_props$patient_id.x[sample_clust$order]

masstag_pheno <- masstag_pheno %>%
  ungroup() %>%
  mutate(patient_id.x = factor(patient_id.x, levels = sample_order))

ggplot(masstag_pheno) +
  geom_tile(
    aes(x = cell_cluster, y = patient_id.x, fill = sqrt(prop))
  ) +
  facet_grid(Clinical.Subtype ~ cell_type, scale = "free", space = "free") +
  scale_fill_gradient(low = "white", high = "black") +
  theme(
    axis.text.x = element_text(angle = 90),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.spacing = unit(0, "cm")
  )
```

## Dimensionality Reduction

* Run a dimensionality reduction algo on the matrix
* Plot the resulting map for a couple of samples
* Overlay the cell types or certain CD's on them
* Write a function to count the proportions / amounts of cells in different
  parts of the map

```{r}

```


## Reduction vs. Phenotypes

* Run a lasso between proportions in different parts of the map and the observed
  phenotypes (say, survival, or cancer grade, or some combination of those)

## Repeat everything on the other dataset

## Relating approaches

* Do you observe similar clusters among better / worse outcome patients?
  - If so, then gives some basis for performing clustering on MIBI that uses
    spatial info, and imputing that spatial info onto the Cytof
* Can you try to "reverse" the lasso from above? From a fixed phenotypic
  configuration, come up with a distribution over plausible maps, for each of
  the two types of data.
